Interpretable Machine Learning for Science \\ with PySR and SymbolicRegression.jl

### 2023-05-02 src/tex/ms.tex
Macro expansions

This brings us to an optimization problem known ``symbolic regression,'' or SR.
In this paper, we describe an algorithm for performing \sr, and describe software dedicated to equation discovery in the sciences.

\paragraph{Symbolic Regression.}
\Sr describes a supervised learning task where the model space is spanned by analytic expressions.
This is commonly solved in a multi-objective optimization framework, jointly minimizing prediction error and model complexity.
In this family of algorithms, instead of fitting concrete parameters in some overparameterized general model, one searches the space of simple analytic expressions for accurate and interpretable models.
In the history of science, scientists have often performed \sr ``manually,'' using a combination of their intuition and trial-and-error to find simple and accurate empirical expressions.
These empirical expressions then might lead to new theoretical insights, such as the aforementioned discoveries of Kepler and Planck leading to classical and quantum mechanics respectively.
\Sr algorithms automate this discovery of empirical equations, exploiting the power of modern computing to test many more expressions than our intuition alone could sort through.

\Sr saw early developments 
as a tool for science beginning in the 1970s and 1980s, with works such as~\cite{gerwinInformationProcessingData1974}, and the creation of the tool \textsc{Bacon} in~\cite{langleyBACONProductionSystem1977} with further development in, e.g.,~\cite{langleyRediscoveringPhysicsBACON1979,langleyBACONDiscoveryConservation1981}.
Via a combination of heuristic strategies for brute force searching over possible expression trees, \textsc{Bacon} and its follow-ups such as \textsc{Fahrenheit}~\cite{langleyDatadrivenApproachesEmpirical1989}, demonstrated the ability to discovery a variety of simple physics laws from idealized data.
The use of genetic algorithms, which allow for a more flexible search space and fewer prior assumptions about these expressions, was popularized with the work of~\cite{kozaGeneticProgrammingMeans1994}.
Finally, with the development of the user-friendly tools \eureqa in~\cite{bongardCoverAutomatedReverse2007,schmidtDistillingFreeFormNatural2009,schmidtSymbolicRegressionImplicit2010}, as well as HeuristicLab in~\cite{wagnerHeuristicLabGenericExtensible2005}, \sr started to become practical for use on real-world and noisy scientific data and empirical equations.

\paragraph{Equation Discovery for Science.} 
Today \eureqa is a closed-source proprietary tool, and has fallen out of common use in the sciences.
In addition to other early work such as~\cite{debFastElitistNondominated2000,debFastElitistMultiobjective2002,davidsonSymbolicNumericalRegression2003,cranmerPhysicsGPGeneticProgramming2005},
there have recently been a significant number of exciting new developments in \sr as well as \sr for science.
These have largely been focused on genetic algorithms on expressions represented as trees, and include core advances to genetic algorithms, such as~\cite{lacavaEpsilonLexicaseSelectionRegression2016,lacavaProbabilisticMultiobjectiveAnalysis2018}, which propose modified strategies for selecting an individual to mutate and~\cite{virgolinScalableGeneticProgramming2017,virgolinImprovingModelbasedGenetic2021} which use an adaptive form of recombination based on emerging statistical patterns.
New strategies for \sr altogether have also been developed:
in particular, our previous work~\cite{cranmerLearningSymbolicPhysics2019,cranmerDiscoveringSymbolicModels2020,cranmerDisentangledSparsityNetworks2021} proposes \sr as a way of interpreting a neural network.
Neural networks can efficiently find low-dimensional patterns in high-dimensional data, and so this mixed strategy also presents a way of performing \sr on high-dimensional or non-tabular data.

Additional techniques include~\cite{jinBayesianSymbolicRegression2020,guimeraBayesianMachineScientist2020} which propose a Markov chain Monte Carlo-like sampling for \sr;
\cite{petersenDeepSymbolicRegression2021,liNeuralguidedSymbolicRegression2019} which trains deep generative models to propose mutated expressions;
\cite{dascoliDeepSymbolicRegression2022,kamiennyEndtoendSymbolicRegression2022} which pre-train a large transformer model on billions of synthetic examples, in order to autoregressively generate expressions from data;
\cite{udrescuAIFeynmanParetooptimal2020,liuAIPoincareMachine2020,wetzelDiscoveringSymmetryInvariants2020} which develop techniques for discovering symmetries with \sr;
\cite{sahooLearningEquationsExtrapolation2018} which optimizes a sparse set of coefficients in a very large pre-defined expression with stochastic gradient descent; and finally~\cite{atkinsonDatadrivenDiscoveryFreeform2019,rossBenchmarkingMachineLearning2022} which apply genetic algorithms to a flexible space of differential equations.

@@

However, here we argue that many existing algorithms lack certain features which make them practical in a broad setting for discovering interpretable equations in science.
The requirements for ``scientific \sr'' are significantly more difficult than for \sr applied to synthetic datasets, which many algorithms are implicitly trained on via synthetic benchmarks.
We list these below.

\begin{easylist}
\ListProperties(Margin1=1cm,Hang1=true,Indent1=.5cm,Numbers1=a,Start1=1,Align=0.5cm)
#
an empirical equation in science quite often contains unknown real constants\footnote{One could argue that any constant in a scientific model must depend on a finite number of known constants of physics (e.g., the speed of light) and mathematics (e.g., $\pi$). However, this introduces significantly more complexity than is being reduced, as the relationship with fundamental constants to derived constants might be arbitrarily complex.}
The set of possible expressions considered by some \sr algorithms such as~\cite{udrescuAIFeynmanParetooptimal2020}: operators, integers, and variables---is discrete and finite up to a given max size.
This set is even searchable with brute force, as is used in~\cite{udrescuAIFeynmanParetooptimal2020} in combination with a symmetry and dimensional analysis module.
However, with arbitrary real constants, the set of equations one must search is uncountably infinite.
The difficulty of searching this set is further compounded by several factors:
#
The discovered equation must provide insight into the nature of the problem.
The equation which often holds the most insight, and is therefore adopted by scientists, is not always the most accurate (many other ML algorithms could be used instead for accuracy), but is instead an equation which balances accuracy with simplicity.
#
An empirical equation in science is discovered from data containing noise, which often has heteroscedastic structure.
#
Many empirical equations are not differentiable, instead being composed of different expressions active in only parts of parameter space (e.g., $\exp(x)$ for $x<0$ and $x^2+1$ for $x\geq 0$).
#
Discovered expressions in science must satisfy known constraints, e.g., mass conservation.
#
Empirical relations are frequently comprised of operators which are unique to one particular field of science, so the equation search strategy must allow for custom operators.
#
Observational data is often high-dimensional, and the \sr algorithm must either manage this internally, or be layered with a feature selection algorithm.
#
The search tool must be capable of finding relationships in non-tabular datasets, such as sequences, grids, and graphs (for example, by using aggregation operators such as $\sum$). #
Finally, and most importantly, any \sr package useful to scientists must be open-source, cross-platform, easy-to-use, and interface with common tools.
For more discussion, the references~\cite{price-whelanAstropyProjectBuilding2018,astropycollaborationAstropyProjectSustaining2022} consider this from the point of view of the astrophysics community.
\end{easylist}

@@

\pysr is a multi-population evolutionary algorithm with multiple evolutions performed asynchronously.
The main loop of \pysr, operating on each population independently, is a classic evolutionary algorithm based on tournament selection for individual selection (first introduced in the thesis of~\cite{brindleGeneticAlgorithmsFunction1980} and generalized in~\cite{goldbergComparativeAnalysisSelection1991}),
and several mutations and crossovers for generating new individuals.
Evolutionary algorithms for \sr was first popularized in~\cite{kozaGeneticProgrammingMeans1994}, and has dominated the development of flexible \sr algorithms since.
\pysr makes several modifications to this classic approach, some of which are motivated by results in recent work.

@@

The second modification we make to tournament selection is specific to the individuals we wish to consider: mathematical expressions.
We embed the genetic algorithm inside an \textit{evolve-simplify-optimize loop}.
``Evolve'' is a repeated application of tournament selection-based evolution for a set number of mutations.
In other words, this is an entire evolution of a population.
``Simplify'' here refers to equation simplification---equations are simplified to an equivalent form during the loop using a set of algebraic equivalencies (since this happens infrequently enough, it tends to not harm discovery by constraining the search space).
The ``Optimize'' part of this algorithm is a few iterations of a classical optimization algorithm (BFGS~\cite{broydenConvergenceClassDoublerank1970} is used by default, though any optimizer in the \texttt{Optim.jl} package can be used~\cite{mogensenOptimMathematicalOptimization2018}) to optimize constants in every equation explicitly.
This part of the algorithm significantly improves the discovery of equations containing real constants, which is essential for practical application of \sr to science.
It has been shown since at least~\cite{topchyFasterGeneticProgramming2001} that performing local gradient searches on numerical constants in expressions can vastly improve the performance of \sr---we closely integrate an optimized and stochastic version of this.

@@

## Thirdly, the user can define a custom likelihood to optimize, which can also take into account weights. This is explained in the following point.
# \textit{Custom losses.}
\pysr supports arbitrary user-defined loss functions, passed as either a string or a function object.
These take a scalar prediction and target, and optionally a scalar weight, and should return a real number greater than 0.
These per-point losses are then summed over the dataset.
With this, the user is capable of defining arbitrary regression objectives, custom likelihoods (defined as log-likelihood of a single data point), and classification-based losses.
This also allows one to define implicit objectives.
# \textit{Custom operators.}
One of the most powerful features of \pysr is the ability to define custom operators.
Different domains of science have functions unique to their field which appear in many formulae and hold a specific meaning.
It is therefore very important that these operators be available in an \sr library, and so \pysr allows custom user-defined unary or binary functions.
So long as a function can be defined as either
$f:\mathbb{R}\rightarrow \mathbb{R}$ or $\mathbb{R}^2\rightarrow\mathbb{R}$, it can be used as a user-defined operator.
\pysr \ul{does not treat built-in operators any differently than user-defined ones.}
Apart from simplification strategies, \pysr does not know the difference between the $+$ operator and a Bessel function.
# \textit{Feature selection.}
Similar to the Gaussian process preprocessing step for denoising, \pysr also uses a simple dimensionality reduction preprocessing step.
Given a user-defined number of features to select, \pysr uses a gradient-boosting tree algorithm to first fit the dataset, then select the most important features.
These features are fed into the main search loop.
# \textit{Constraints.}
\pysr allows various hard constraints to be specified for discovered expressions.
These are enforced at every mutation: if a constraint is violated, the mutation is rejected. A few are available with the API:
## The maximum size of an expression (the number of instances of operators, variables, and constants), which gives an overall bound on the search space.
## The maximum depth of an expression, which can be used to control how deeply nested the resultant expression is.
## The maximum size of a subexpression in a specific operator.
This specifies the maximum size of an expression within an operator's arguments.
For example, specifying that the $\wedge$ operator has maximum argument size of $(-1, 3)$ means that its base can have any size expression (indicated by -1), while the expression in its exponent can only have max size of up to 3.
This operator-specific constraint can drastically reduce the complexity of discovered expressions.
## The maximum number of nests of a particular operator combination. For example, the nested constraint \texttt{\{sin: \{sin: 0, cos: 1\}\}} would indicate that $\sin(\cos(x))$ is allowed, but $\sin(\cos(\cos(x)))$ and $\sin(\sin(x))$ are not.
# \textit{Additional data structures}. See deep learning interface \cref{sec:interfaces}.
# \textit{Integrals and dynamical systems}. See deep learning interface \cref{sec:interfaces}.
\end{easylist}

@@

The occurrence rate of a particular mathematical operator in a set of models varies by the scientific field those models describe.
For example, models in Epidemiology commonly use exponentials, indicating exponential growth or decay of a contagious disease, but may rarely use a Bessel function — which would be more common in in classical physics applications.
The entire reason that \sr produces interpretable models is because the generated expressions make use of operators which are common in a particular field, and a domain scientist may be able to see connections with existing models.
In some ways, this process relates to language: black box machine learning models represent functions in a language uninterpretable to a scientist, whereas \sr uses the language in place.
Furthermore, due to the modular nature of scientific modelling---many complex models are built on-top of existing simple models for particular subsystems---it also makes sense from a standpoint of improving model accuracy to make use of common operators (for an interesting explicit example of this, see~\cite{guimeraBayesianMachineScientist2020}).

@@

In many machine learning toy datasets for benchmarking regression algorithms, Mean-Square-Error (MSE or $L_2$ loss) is typically used as a learning objective~\cite{grinsztajnWhyTreebasedModels2022}.
In a Bayesian framework, MSE is equivalent to assuming every data point is Gaussian distributed, with equal variance per point.
Minimizing MSE is equivalent to maximizing the Gaussian log-likelihood.
However, in science, one typically works with 
a likelihood that is very specific to a particular problem,
and this is often non-Gaussian.
Therefore, it is important for an \sr package to allow for custom loss functions.
\pysr implements this in a way that is very similar to that of custom operators (see \cref{sec:custom_operators}).
Given a string such as ``\texttt{loss(x, y) = abs(x - y)}'', \pysr will pass this to the Julia backend, which will automatically vectorize it and use it as a loss function throughout the search process.
In a Bayesian context, this would allow one to define arbitrary likelihoods, even for very complex branching logic.
This also works for weighted losses, such as ``\texttt{loss(x, y, w) = abs(x - y) * w}''

@@

First, we present a high level comparison of the various \sr tools available.
These qualitative points are made using the desirable features given in \cref{sec:intro_pysr}.

@@

Existing \sr benchmarks have certain simplifications compared to the datasets used to historically discover well-known empirical equations.
When discovering a new expression, one does not actually know the physical constants in the expression, and one must have to learn real constants.
For example, in the Feynman benchmark dataset of~\cite{udrescuAIFeynmanPhysicsinspired2020}, all expressions are listed with the associated physical constants---but these constants originally had to be discovered along with the equation.
Similarly, in the commonly-used synthetic ``Nguyen'' benchmark~\cite{nguyenquanguySemanticallybasedCrossoverGenetic2011}, which includes expressions such as $F_9=\sin(x)+\sin(y^2)$, there are no non-integral constants.
The ``SRBench'' competition of~\cite{defrancaInterpretableSymbolicRegression2023} was meticulous in its comparisons of different methods, but only contained a single real-world task (with an unknown ground truth), with all other expressions being synthetic.
This benchmark improved on others in that it included tunable amounts of noise.
However, the noise models are very simple: perfectly Gaussian and including no heteroscedasticity.
Although such synthetic benchmarks have their uses in \sr research, these types of benchmarks are often not entirely indicative of the challenges faced in real-world scientific discovery.
Therefore, here we introduce a new benchmark which attempts to accurately portray the empirical equation discovery step of science.
Every equation in this dataset is a real empirical expression that a scientist has at one point discovered from experimental, noisy, and imperfect data.

@@

\begin{figure*}[h!]
    \script{equations.py}
    \begin{center}
        \includegraphics[width=0.95\linewidth]{figures/benchmark_data.pdf}
    \end{center}
    \caption{Visualization of all the data in \bench, an \sr benchmark for science.
    Color is used to denote additional variables in the cases of relations which depend on more than two inputs.
    Original data in the discovery of each law is used where easily available.
    Otherwise, data is generated from the formula with realistic ranges of variables, with a level of noise applied.
    }
    \label{fig:bench_data}
\end{figure*}

@@

Many existing benchmarks for physical expression searches are based on theoretically-derived equations rather than empirically-observed and formulated. 
While these are likely useful for benchmarking some parts of the \sr landscape, they do not target the aspect which \sr would ultimately be used for: empirical discovery.
Thus, all the equations in this dataset were first empirically formulated before (and if) they were theoretically-derived.
Thus, the physical variables here directly correspond to observables.
Furthermore, this dataset does not include any relevant physical constants---the algorithm must find these automatically, as did the scientist who discovered the equation.

@@

We emphasize that the stability of an \sr algorithm and software implementation itself has central importance to practical use in science, and therefore, unstable runs which either produce an undefined expression, or crash, are \textit{included} in the results dataset and the primary score.
However, one should also note that because we are the authors of \pysr, we are therefore more likely to be running in a stable environment than, say, a package which we are unfamiliar with.
For example, as seen in \cref{tab:results}, DSR sees improved results if one ignores the failed runs, although there were still some expressions it struggled on.
Thus, detailed itemization of these results is given in \cref{tab:results}, to allow for a more nuanced understanding of the results.
Also note that this benchmark does not explicitly measure other metrics which are considered in other benchmarks, such as number of evaluations (where, e.g., genetic algorithms are very inefficient), nor does it consider the numerical accuracy of the expressions.

@@

What is also interesting is that the two pure deep learning approaches, EQL and SR-Transformer, recovered the fewest expressions out of all tested algorithms.
One of these, EQL, learns the expression in an online fashion, while SR-Transformer performs fast inference based on pre-trained weights.
SR-Transformer, in particular, is pre-trained on billions of randomly-generated synthetic expressions on a cluster of GPUs for weeks.
However, the ``untrained'' algorithms written using classic heuristics: PySR, Operon, DSR, and QLattice, all out-performed these modern deep learning models. 
Perhaps this is an insight into the difficulty of preparing deep learning systems for real-world data and unexpected ``long-tail'' examples, whereas handwritten standard algorithms will often perform just as well on unexpected examples.  In \sr especially, the space of potential expressions is so massive and nonlinear that interpolation techniques might suffer; consider that $x/y$ and $x\times y$ are just a single mutation away, but infinitely distinct in their data representation.
In spite of this, while pure deep learning strategies appears to require improvements in architecture or training to be useful in a real world \sr setting, its success in, e.g.,~\cite{cranmerDiscoveringSymbolicModels2020,petersenDeepSymbolicRegression2021,kamiennyEndtoendSymbolicRegression2022} shows there is indeed strong potential long-term for such hybrid methods, and these directions should continue to be pursued.

>>>>

This brings us to an optimization problem known ``symbolic regression,'' or SR.
In this paper, we describe an algorithm for performing SR, and describe software dedicated to equation discovery in the sciences.

\paragraph{Symbolic Regression.}
SR describes a supervised learning task where the model space is spanned by analytic expressions.
This is commonly solved in a multi-objective optimization framework, jointly minimizing prediction error and model complexity.
In this family of algorithms, instead of fitting concrete parameters in some overparameterized general model, one searches the space of simple analytic expressions for accurate and interpretable models.
In the history of science, scientists have often performed SR ``manually,'' using a combination of their intuition and trial-and-error to find simple and accurate empirical expressions.
These empirical expressions then might lead to new theoretical insights, such as the aforementioned discoveries of Kepler and Planck leading to classical and quantum mechanics respectively.
SR algorithms automate this discovery of empirical equations, exploiting the power of modern computing to test many more expressions than our intuition alone could sort through.

SR saw early developments 
as a tool for science beginning in the 1970s and 1980s, with works such as~\cite{gerwinInformationProcessingData1974}, and the creation of the tool \textsc{Bacon} in~\cite{langleyBACONProductionSystem1977} with further development in, e.g.,~\cite{langleyRediscoveringPhysicsBACON1979,langleyBACONDiscoveryConservation1981}.
Via a combination of heuristic strategies for brute force searching over possible expression trees, \textsc{Bacon} and its follow-ups such as \textsc{Fahrenheit}~\cite{langleyDatadrivenApproachesEmpirical1989}, demonstrated the ability to discovery a variety of simple physics laws from idealized data.
The use of genetic algorithms, which allow for a more flexible search space and fewer prior assumptions about these expressions, was popularized with the work of~\cite{kozaGeneticProgrammingMeans1994}.
Finally, with the development of the user-friendly tools \eureqa in~\cite{bongardCoverAutomatedReverse2007,schmidtDistillingFreeFormNatural2009,schmidtSymbolicRegressionImplicit2010}, as well as HeuristicLab in~\cite{wagnerHeuristicLabGenericExtensible2005}, SR started to become practical for use on real-world and noisy scientific data and empirical equations.

\paragraph{Equation Discovery for Science.} 
Today \eureqa is a closed-source proprietary tool, and has fallen out of common use in the sciences.
In addition to other early work such as~\cite{debFastElitistNondominated2000,debFastElitistMultiobjective2002,davidsonSymbolicNumericalRegression2003,cranmerPhysicsGPGeneticProgramming2005},
there have recently been a significant number of exciting new developments in SR as well as SR for science.
These have largely been focused on genetic algorithms on expressions represented as trees, and include core advances to genetic algorithms, such as~\cite{lacavaEpsilonLexicaseSelectionRegression2016,lacavaProbabilisticMultiobjectiveAnalysis2018}, which propose modified strategies for selecting an individual to mutate and~\cite{virgolinScalableGeneticProgramming2017,virgolinImprovingModelbasedGenetic2021} which use an adaptive form of recombination based on emerging statistical patterns.
New strategies for SR altogether have also been developed:
in particular, our previous work~\cite{cranmerLearningSymbolicPhysics2019,cranmerDiscoveringSymbolicModels2020,cranmerDisentangledSparsityNetworks2021} proposes SR as a way of interpreting a neural network.
Neural networks can efficiently find low-dimensional patterns in high-dimensional data, and so this mixed strategy also presents a way of performing SR on high-dimensional or non-tabular data.

Additional techniques include~\cite{jinBayesianSymbolicRegression2020,guimeraBayesianMachineScientist2020} which propose a Markov chain Monte Carlo-like sampling for SR;
\cite{petersenDeepSymbolicRegression2021,liNeuralguidedSymbolicRegression2019} which trains deep generative models to propose mutated expressions;
\cite{dascoliDeepSymbolicRegression2022,kamiennyEndtoendSymbolicRegression2022} which pre-train a large transformer model on billions of synthetic examples, in order to autoregressively generate expressions from data;
\cite{udrescuAIFeynmanParetooptimal2020,liuAIPoincareMachine2020,wetzelDiscoveringSymmetryInvariants2020} which develop techniques for discovering symmetries with SR;
\cite{sahooLearningEquationsExtrapolation2018} which optimizes a sparse set of coefficients in a very large pre-defined expression with stochastic gradient descent; and finally~\cite{atkinsonDatadrivenDiscoveryFreeform2019,rossBenchmarkingMachineLearning2022} which apply genetic algorithms to a flexible space of differential equations.

@@

However, here we argue that many existing algorithms lack certain features which make them practical in a broad setting for discovering interpretable equations in science.
The requirements for ``scientific SR'' are significantly more difficult than for SR applied to synthetic datasets, which many algorithms are implicitly trained on via synthetic benchmarks.
We list these below.

\begin{easylist}
\ListProperties(Margin1=1cm,Hang1=true,Indent1=.5cm,Numbers1=a,Start1=1,Align=0.5cm)
#
an empirical equation in science quite often contains unknown real constants\footnote{One could argue that any constant in a scientific model must depend on a finite number of known constants of physics (e.g., the speed of light) and mathematics (e.g., $\pi$). However, this introduces significantly more complexity than is being reduced, as the relationship with fundamental constants to derived constants might be arbitrarily complex.}
The set of possible expressions considered by some SR algorithms such as~\cite{udrescuAIFeynmanParetooptimal2020}: operators, integers, and variables---is discrete and finite up to a given max size.
This set is even searchable with brute force, as is used in~\cite{udrescuAIFeynmanParetooptimal2020} in combination with a symmetry and dimensional analysis module.
However, with arbitrary real constants, the set of equations one must search is uncountably infinite.
The difficulty of searching this set is further compounded by several factors:
#
The discovered equation must provide insight into the nature of the problem.
The equation which often holds the most insight, and is therefore adopted by scientists, is not always the most accurate (many other ML algorithms could be used instead for accuracy), but is instead an equation which balances accuracy with simplicity.
#
An empirical equation in science is discovered from data containing noise, which often has heteroscedastic structure.
#
Many empirical equations are not differentiable, instead being composed of different expressions active in only parts of parameter space (e.g., $\exp(x)$ for $x<0$ and $x^2+1$ for $x\geq 0$).
#
Discovered expressions in science must satisfy known constraints, e.g., mass conservation.
#
Empirical relations are frequently comprised of operators which are unique to one particular field of science, so the equation search strategy must allow for custom operators.
#
Observational data is often high-dimensional, and the SR algorithm must either manage this internally, or be layered with a feature selection algorithm.
#
The search tool must be capable of finding relationships in non-tabular datasets, such as sequences, grids, and graphs (for example, by using aggregation operators such as $\sum$). #
Finally, and most importantly, any SR package useful to scientists must be open-source, cross-platform, easy-to-use, and interface with common tools.
For more discussion, the references~\cite{price-whelanAstropyProjectBuilding2018,astropycollaborationAstropyProjectSustaining2022} consider this from the point of view of the astrophysics community.
\end{easylist}

@@

\pysr is a multi-population evolutionary algorithm with multiple evolutions performed asynchronously.
The main loop of \pysr, operating on each population independently, is a classic evolutionary algorithm based on tournament selection for individual selection (first introduced in the thesis of~\cite{brindleGeneticAlgorithmsFunction1980} and generalized in~\cite{goldbergComparativeAnalysisSelection1991}),
and several mutations and crossovers for generating new individuals.
Evolutionary algorithms for SR was first popularized in~\cite{kozaGeneticProgrammingMeans1994}, and has dominated the development of flexible SR algorithms since.
\pysr makes several modifications to this classic approach, some of which are motivated by results in recent work.

@@

The second modification we make to tournament selection is specific to the individuals we wish to consider: mathematical expressions.
We embed the genetic algorithm inside an \textit{evolve-simplify-optimize loop}.
``Evolve'' is a repeated application of tournament selection-based evolution for a set number of mutations.
In other words, this is an entire evolution of a population.
``Simplify'' here refers to equation simplification---equations are simplified to an equivalent form during the loop using a set of algebraic equivalencies (since this happens infrequently enough, it tends to not harm discovery by constraining the search space).
The ``Optimize'' part of this algorithm is a few iterations of a classical optimization algorithm (BFGS~\cite{broydenConvergenceClassDoublerank1970} is used by default, though any optimizer in the \texttt{Optim.jl} package can be used~\cite{mogensenOptimMathematicalOptimization2018}) to optimize constants in every equation explicitly.
This part of the algorithm significantly improves the discovery of equations containing real constants, which is essential for practical application of SR to science.
It has been shown since at least~\cite{topchyFasterGeneticProgramming2001} that performing local gradient searches on numerical constants in expressions can vastly improve the performance of SR---we closely integrate an optimized and stochastic version of this.

@@

## Thirdly, the user can define a custom likelihood to optimize, which can also take into account weights. This is explained in the following point.
# \textit{Custom losses.}
\pysr supports arbitrary user-defined loss functions, passed as either a string or a function object.
These take a scalar prediction and target, and optionally a scalar weight, and should return a real number greater than 0.
These per-point losses are then summed over the dataset.
With this, the user is capable of defining arbitrary regression objectives, custom likelihoods (defined as log-likelihood of a single data point), and classification-based losses.
This also allows one to define implicit objectives.
# \textit{Custom operators.}
One of the most powerful features of \pysr is the ability to define custom operators.
Different domains of science have functions unique to their field which appear in many formulae and hold a specific meaning.
It is therefore very important that these operators be available in an SR library, and so \pysr allows custom user-defined unary or binary functions.
So long as a function can be defined as either
$f:\mathbb{R}\rightarrow \mathbb{R}$ or $\mathbb{R}^2\rightarrow\mathbb{R}$, it can be used as a user-defined operator.
\pysr \ul{does not treat built-in operators any differently than user-defined ones.}
Apart from simplification strategies, \pysr does not know the difference between the $+$ operator and a Bessel function.
# \textit{Feature selection.}
Similar to the Gaussian process preprocessing step for denoising, \pysr also uses a simple dimensionality reduction preprocessing step.
Given a user-defined number of features to select, \pysr uses a gradient-boosting tree algorithm to first fit the dataset, then select the most important features.
These features are fed into the main search loop.
# \textit{Constraints.}
\pysr allows various hard constraints to be specified for discovered expressions.
These are enforced at every mutation: if a constraint is violated, the mutation is rejected. A few are available with the API:
## The maximum size of an expression (the number of instances of operators, variables, and constants), which gives an overall bound on the search space.
## The maximum depth of an expression, which can be used to control how deeply nested the resultant expression is.
## The maximum size of a subexpression in a specific operator.
This specifies the maximum size of an expression within an operator's arguments.
For example, specifying that the $\wedge$ operator has maximum argument size of $(-1, 3)$ means that its base can have any size expression (indicated by -1), while the expression in its exponent can only have max size of up to 3.
This operator-specific constraint can drastically reduce the complexity of discovered expressions.
## The maximum number of nests of a particular operator combination. For example, the nested constraint \texttt{\{sin: \{sin: 0, cos: 1\}\}} would indicate that $\sin(\cos(x))$ is allowed, but $\sin(\cos(\cos(x)))$ and $\sin(\sin(x))$ are not.
# \textit{Additional data structures}. See deep learning interface \cref{sec:interfaces}.
# \textit{Integrals and dynamical systems}. See deep learning interface \cref{sec:interfaces}.
\end{easylist}

@@

The occurrence rate of a particular mathematical operator in a set of models varies by the scientific field those models describe.
For example, models in Epidemiology commonly use exponentials, indicating exponential growth or decay of a contagious disease, but may rarely use a Bessel function — which would be more common in in classical physics applications.
The entire reason that SR produces interpretable models is that the generated expressions make use of operators which are common in a particular field, and a domain scientist may be able to see connections with existing models.
In some ways, this process relates to language: black box machine learning models represent functions in a language uninterpretable to a scientist, whereas SR uses the language in place.
Furthermore, due to the modular nature of scientific modelling---many complex models are built on-top of existing simple models for particular subsystems---it also makes sense from a standpoint of improving model accuracy to make use of common operators (for an interesting explicit example of this, see~\cite{guimeraBayesianMachineScientist2020}).

@@

In many machine learning toy datasets for benchmarking regression algorithms, Mean-Square-Error (MSE or $L_2$ loss) is typically used as a learning objective~\cite{grinsztajnWhyTreebasedModels2022}.
In a Bayesian framework, MSE is equivalent to assuming every data point is Gaussian distributed, with equal variance per point.
Minimizing MSE is equivalent to maximizing the Gaussian log-likelihood.
However, in science, one typically works with 
a likelihood that is very specific to a particular problem,
and this is often non-Gaussian.
Therefore, it is important for an SR package to allow for custom loss functions.
\pysr implements this in a way that is very similar to that of custom operators (see \cref{sec:custom_operators}).
Given a string such as ``\texttt{loss(x, y) = abs(x - y)}'', \pysr will pass this to the Julia backend, which will automatically vectorize it and use it as a loss function throughout the search process.
In a Bayesian context, this would allow one to define arbitrary likelihoods, even for very complex branching logic.
This also works for weighted losses, such as ``\texttt{loss(x, y, w) = abs(x - y) * w}''

@@

First, we present a high level comparison of the various SR tools available.
These qualitative points are made using the desirable features given in \cref{sec:intro_pysr}.

@@

Existing SR benchmarks have certain simplifications compared to the datasets used to historically discover well-known empirical equations.
When discovering a new expression, one does not actually know the physical constants in the expression, and one must have to learn real constants.
For example, in the Feynman benchmark dataset of~\cite{udrescuAIFeynmanPhysicsinspired2020}, all expressions are listed with the associated physical constants---but these constants originally had to be discovered along with the equation.
Similarly, in the commonly-used synthetic ``Nguyen'' benchmark~\cite{nguyenquanguySemanticallybasedCrossoverGenetic2011}, which includes expressions such as $F_9=\sin(x)+\sin(y^2)$, there are no non-integral constants.
The ``SRBench'' competition of~\cite{defrancaInterpretableSymbolicRegression2023} was meticulous in its comparisons of different methods, but only contained a single real-world task (with an unknown ground truth), with all other expressions being synthetic.
This benchmark improved on others in that it included tunable amounts of noise.
However, the noise models are very simple: perfectly Gaussian and including no heteroscedasticity.
Although such synthetic benchmarks have their uses in SR research, these types of benchmarks are often not entirely indicative of the challenges faced in real-world scientific discovery.
Therefore, here we introduce a new benchmark which attempts to accurately portray the empirical equation discovery step of science.
Every equation in this dataset is a real empirical expression that a scientist has at one point discovered from experimental, noisy, and imperfect data.

@@

\begin{figure*}[h!]
    \script{equations.py}
    \begin{center}
        \includegraphics[width=0.95\linewidth]{figures/benchmark_data.pdf}
    \end{center}
    \caption{Visualization of all the data in \bench, an SR benchmark for science.
    Color is used to denote additional variables in the cases of relations which depend on more than two inputs.
    Original data in the discovery of each law is used where easily available.
    Otherwise, data is generated from the formula with realistic ranges of variables, with a level of noise applied.
    }
    \label{fig:bench_data}
\end{figure*}

@@

Many existing benchmarks for physical expression searches are based on theoretically-derived equations rather than empirically-observed and formulated. 
While these are likely useful for benchmarking some parts of the SR landscape, they do not target the aspect which SR would ultimately be used for: empirical discovery.
Thus, all the equations in this dataset were first empirically formulated before (and if) they were theoretically-derived.
Thus, the physical variables here directly correspond to observables.
Furthermore, this dataset does not include any relevant physical constants---the algorithm must find these automatically, as did the scientist who discovered the equation.

@@

We emphasize that the stability of an SR algorithm and software implementation itself has central importance to practical use in science, and therefore, unstable runs which either produce an undefined expression, or crash, are \textit{included} in the results dataset and the primary score.
However, one should also note that because we are the authors of \pysr, we are therefore more likely to be running in a stable environment than, say, a package which we are unfamiliar with.
For example, as seen in \cref{tab:results}, DSR sees improved results if one ignores the failed runs, although there were still some expressions it struggled on.
Thus, detailed itemization of these results is given in \cref{tab:results}, to allow for a more nuanced understanding of the results.
Also note that this benchmark does not explicitly measure other metrics which are considered in other benchmarks, such as number of evaluations (where, e.g., genetic algorithms are very inefficient), nor does it consider the numerical accuracy of the expressions.

@@

What is also interesting is that the two pure deep learning approaches, EQL and SR-Transformer, recovered the fewest expressions out of all tested algorithms.
One of these, EQL, learns the expression in an online fashion, while SR-Transformer performs fast inference based on pre-trained weights.
SR-Transformer, in particular, is pre-trained on billions of randomly-generated synthetic expressions on a cluster of GPUs for weeks.
However, the ``untrained'' algorithms written using classic heuristics: PySR, Operon, DSR, and QLattice, all out-performed these modern deep learning models. 
Perhaps this is an insight into the difficulty of preparing deep learning systems for real-world data and unexpected ``long-tail'' examples, whereas handwritten standard algorithms will often perform just as well on unexpected examples.  In SR especially, the space of potential expressions is so massive and nonlinear that interpolation techniques might suffer; consider that $x/y$ and $x\times y$ are just a single mutation away, but infinitely distinct in their data representation.
In spite of this, while pure deep learning strategies appears to require improvements in architecture or training to be useful in a real world SR setting, its success in, e.g.,~\cite{cranmerDiscoveringSymbolicModels2020,petersenDeepSymbolicRegression2021,kamiennyEndtoendSymbolicRegression2022} shows there is indeed strong potential long-term for such hybrid methods, and these directions should continue to be pursued.
